\chapter{Standard libraries}

% -----------------
% SECTION
% -----------------
\section{Runtime libraries}

A program may request libraries as shared objects,
if the shared object name has no slash then the loader find for it
first on directories specified by the environment variable
\texttt{LD\_LIBRARY\_PATH} and then in \texttt{<TUIM\_HOME>/lib}.

All shared object names are reserved by this ABI,
if a  need a shared object for porposes others than
specified in this document then it need provide the shared object and
the environment variable \texttt{LD\_LIBRARY\_PATH} need to be set
after program loading.

All global and weak symbols prefixed with \texttt{\_\_tuim\_}
are reserved by this ABI,
a conforming program shall not provide definitions for these symbols,
even if it not request any library.

% -----------------
% SECTION
% -----------------
\section{Compiler runtime library}

The \texttt{libcrt.so} shared object
expose symbols from the Compiler Runtime Library,
usually helper functions that the compiler need.
Symbols exposed by this shared object
begins with two underscores or an underscore followed by a capital letter.

Some processor suplements provide a specification for the
Compiler Runtime Library, in that case this is the library in the shared
object \texttt{libcrt.so}.
If the processor suplement do not provide such specification than
is a undefined bahavior the explicit use of
any symbol exposed by \texttt{libcrt.so} in source code,
these symbols are reserved for the use of the compiler.

TODO: Alternatively, I want to provide a CLang compatible ABI for processors
without such specification.

% -----------------
% SECTION
% -----------------
\section{The C standard library}

The \texttt{libc.so} shared object
expose symbols from the C Standard Library, the latest version\cite{ISO9899}.
Symbols exposed by this shared object are symbols reserved by the C standard.

Several global and weak symbols
are reserved by the C standard,
a program requesting \texttt{libc.so}
shall not provide definitions for these symbols.
These are\footnote{%
   Note that the nomeclature used by the C standard is
   "identifier" instead of "symbol" as used by the ELF standard,
   however all external identifiers are mapped by the compiler in
   global or weak symbols.
   For more details please check the C standard\cite{ISO9899}.
}:

\begin{itemize}
   \item Symbols prefixed with an underscore (\texttt{\_}).
   \item Symbols defined by the standard library (in hosted environment).
   \item Symbols declared as reserved for the implementation
   or future use by the standard library.
   \item Symbols declared as potentially reserved.
\end{itemize}
